\documentclass[in]{ucl_document}

\title[\noexpand \LaTeXe\ Macros for Constructing UCL
Documents]{\LaTeXe\ Macros for Constructing UCL Documents}

\author{Dr Russel Winder}

\personname{Dr Russel Winder}
\persontitle{Reader in Software Engineering}
\personrole{ex Research Student Tutor}
\persontelephone{+44 (0)171 380 7293}
\personemail{R.Winder@cs.ucl.ac.uk}
\personurl{//www.cs.ucl.ac.uk/staff/R.Winder/}
\personpostnominal{MInstP, CPhys, FBCS, CEng}
\setboolean{personrolewanted}{false}

\documentnumber{95/XX}

\newcommand \acronym[1]{{\small #1}}

\begin{document}

\maketitle

\begin{abstract}
This document should provide sufficient information for people to use
the \texttt{ucl\_letter}, \texttt{ucl\_fax}, \texttt{ucl\_memo},
\texttt{ucl\_document}, \texttt{ucl\_thesis} and \texttt{ucl\_exam}
classes.

It should be noted that these classes have been used quite extensively
by the author but have not been extensively tested beyond normal use
by the author.  This means that there are likely to be a few bugs
resulting in interesting behaviour in some circumstances of use that
the author does not himself employ.

If you find any such situations or if you find any other errors or
omissions in either the macros or this document, please contact the
author.
\end{abstract}

\section{Introduction}

The Department of Computer Science at \acronym{UCL} has a standard
format for letters, faxes, memos, documents --- internal notes
(\acronym{IN}), research notes (\acronym{RN}) and technical reports
(\acronym{TR}) --- and exam papers.  The standard is implemented in
\LaTeXe, psroff and Word, which are the 3 formatting mechanisms most
popular in the department.  At some stage FrameMaker templates may
also be produced.

This document covers how to use the \LaTeXe\ classes.  Separate
documents cover the other formatting systems.

\section{The Basic Structure}

The \acronym{UCL} macros are not complete classes in themselves but
are extension of the \LaTeXe\ standard classes; the classes inherit
directly from standard ones.  \texttt{ucl\_letter} and
\texttt{ucl\_memo} inherit from \texttt{letter}; \texttt{ucl\_fax}
inherits from \texttt{ucl\_letter} and hence is also a
\texttt{letter}; \texttt{ucl\_thesis} inherits from \texttt{book}; and
\texttt{ucl\_document} and \texttt{ucl\_exam} inherit from
\texttt{article}.  Access to \textit{\LaTeX\ A Document Preparation
System}, 2nd edition by Leslie Lamport (or something equivalent) is
therefore required since this document deals only with the extensions
provided by the \texttt{ucl\_*} classes.

Letter, faxes and documents come in two guises, general and
personalized.  The general guise produces a format with general
departmental information and requires no extra information about the
user.  The personalized forms require the user to have predefined a
package, with certain macro calls inside, included into the document
using \verb+\usepackage+ (or to include the macro calls directly in
the document source, of course) in order to prepare the personal
information in the right form.

In fact, there is a command
\texttt{/opt/tex/lib/texmf/tex/ucl/preparesignature} which will
prepare the basics of such a file from the information stored in
\texttt{/cs/docs/research/deptlist.awk}.  This is the file containing
the official information held about people in the central user
database.

\section{The Personal Signature File}
\label{sec:personalize}

The information required in a personal signature file is best
demonstrated via an example.  The following is the result of running
the \texttt{preparesignature} file for myself:
\begin{quote}
\begin{verbatim}
\personname{Dr Russel Winder}
\persontitle{Reader in Software Engineering}
\personrole{ex Research Student Tutor}
\persontelephone{+44 (0)171 380 7293}
\personemail{R.Winder@cs.ucl.ac.uk}
\personurl{//www.cs.ucl.ac.uk/staff/R.Winder/}
\personpostnominal{MInstP, CPhys, FBCS, CEng}
\end{verbatim}
\end{quote}

By default the classes use all the fields one way or another.  It is
sometimes required that some of the fields are not used.  The
printing, or not, of a field is controlled by a Boolean,
\texttt{person*wanted}, where \texttt{*} is one of \texttt{title},
\texttt{role}, \texttt{telephone}, \texttt{mail}, \texttt{url} and
\texttt{postnominal}.  A line such as:
\begin{quote}
\begin{verbatim}
\setboolean{personrolewanted}{false}
\end{verbatim}
\end{quote}
causes the class not to use the role field.

This information could in principle be included directly in the
\LaTeXe\ source document (as it is in fact with this document) but
more usually it will reside in a file with a \texttt{.sty} extension
and be included into the document as a package, i.e.\ with a line such
as:
\begin{quote}
\begin{verbatim}
\usepackage{rlw_signature}
\end{verbatim}
\end{quote}
in the source file.

\section{UCL Letters, Faxes and Memos}

\acronym{UCL} letters are letters with a special letterhead which may
or may not be personalized.  A start line of:
\begin{quote}
\begin{verbatim}
\documentclass{ucl_letter}
\end{verbatim}
\end{quote}
sets everything up for a general \acronym{UCL} letter, whereas using
the option \texttt{personalisehead} uses the personalized form of the
letterhead:
\begin{quote}
\begin{verbatim}
\documentclass[personalisehead]{ucl_letter}
\end{verbatim}
\end{quote}
Everything else about a letter is as Standard \LaTeXe\ letter class
and environment.

Faxes are no different from letters except that one extra field is
required the fax number to send to.  An extra command is introduce:
\begin{quote}
\begin{verbatim}
\tofax{...}
\end{verbatim}
\end{quote}
which appears in the preamble.

A memo is a letter with a simpler letterhead.  It is accessed using
the class \verb+ucl_memo+.  However, it requires a subject field so a
call to:
\begin{quote}
\begin{verbatim}
\subject{...}
\end{verbatim}
\end{quote}
is required to set the field.

\section{UCL Documents}

\acronym{UCL} documents include \acronym{IN}s, \acronym{RN}s and
\acronym{TR}s which are simply variations on a theme, the
\acronym{UCL} document.  Thus, \texttt{in}, \texttt{rn}, and
\texttt{tr} are simply options to the \texttt{ucl\_document} class.
Thus for example the start line might be:
\begin{quote}
\begin{verbatim}
\documentclass[rn]{ucl_document}
\end{verbatim}
\end{quote}
This class can use the personalization data as described in
Section~\ref{sec:personalize}.  If you use this facility then the
\verb+\author+ macro does not need to be used in the document to get
the name on the front cover.  If you choose not to use this facility
then not only must \verb+\author+ be used but a call of the following
ilke must appear to deal with all contact details:
\begin{quote}
\begin{verbatim}
\contactdetails{%
    \mbox{\emph{Telephone}: +44 (0)171 380 7293}\\%
    \mbox{\emph{Fax}: +44 (0)171 387 1397}\\%
    \mbox{\emph{Electronic Mail}: R.Winder@cs.ucl.ac.uk}\\%
    \mbox{\emph{URL}: http://www.cs.ucl.ac.uk/staff/R,Winder/}\\%
}
\end{verbatim}
\end{quote}

Two extra commands are introduced over and above those required by the
\texttt{article} class and they are:
\begin{quote}
\begin{verbatim}
\documentnumber{...}
\keywords{...}
\end{verbatim}
\end{quote}
\verb+\documentnumber+  should contain the correct document number as
assigned by the department assigner, currently Tessa Rickards, and
\verb+\keywords+ should contain the list of keywords to appear on the
cover of the document.

Apart from the above a \acronym{UCL} document is the same as an
\texttt{article}.

Will except that the \verb+\title+ and \verb+\author+ have been
redefined to admit an optional parameter enabling easy specification
of separate title page title/author and running head title/author.
Thus, these macros can appear like:
\begin{quote}
\begin{verbatim}
\title[Short title]{An Extremely Long Title that is Fine on the Title
Page but would be Useless as a Running Head}
\author[Bloggs et al.][J Bloggs, U Persson, Z Beeblebrox, J
Doe, T Jig}
\end{verbatim}
\end{quote}

Because of the way these redefined versions of \verb+\title+ and
\verb+\author+ work there are minor difficulties using macros in
titles and author lists.  The cure is quite straightforward but
irritating.  I am trying to find a more elegant solution.  An example,
the use of \verb+\title+ for this document, shows the solution, which
is to use the full form of the macro (despite the repetition) and to
disallow early expansion of the macro in the troublesome place:
\begin{quote}
\begin{verbatim}
\title[\noexpand \LaTeXe\ Macros for Constructing UCL
Documents]{\LaTeXe\ Macros for Constructing UCL Documents}
\end{verbatim}
\end{quote}

\section{UCL Theses}

MRes, MPhil and PhD theses have to be typeset to certain rules as laid
down by University of London.  These rules are (more or less)
encapsulated in the \texttt{ucl\_thesis} class, which inherits from
\texttt{book}.

The University regulations on margins, spacing etc.\ are:
\begin{itemize}
\item
Margin at binding edge = 40mm.
\item
Margin at all other edges = 20mm.
\item
Inter-line spacing = 1.5.
\item
Pages must be numbered in one continuous sequence from title page
to the last page of type in arabic numerals.
\end{itemize}

As a thesis is a book, use can be made of \verb+\frontmatter+,
\verb+\mainmatter+ and \verb+\backmatter+ if that is desired but this
would be a variation of the thesis standard since \verb+\frontmatter+
causes roman page numbering and the standard states that numbering is
arabic throughout starting with the title page as page 1.  Despite
this, no thesis has been rejected for being non-standard on this
point, so it is a choice.

Other components of the style are:
\begin{itemize}
\item
\emph{Contents:}  No deeper than subsections.
\item
\emph{Equations, figures, tables and other numbered objects:}  These are
numbered by chapter and item number within chapter.  For example, the
first equation in chapter 1 will be equation 1.1.
\item
\emph{Section headings:}  There is no italic type in chapter, section,
subsection and subsubsection headings.  Bold font at different font
sizes is used.  Subsubsections are not numbered.
\item
\emph{Paragraph style:}  The default should be indented paragraphs with a
supported option for blocked paragraphs.
\item
\emph{Fonts:}  Times Roman is a font that has been consistently used over a
long period of time, and for that reason is preferable to any other
font.  Although the default font is a 10pt one, 11pt font is preferred
(12pt could even be used) so it is usual to put the 11pt option in the
\verb+\documentclass+ statement.
\item
\emph{References:}  References should be all at the end of the thesis and in
alphabetical order.  There is no default on the style of references,
this is left to the author.
\end{itemize}

It is expected that acknowledgements appear between the abstract and
the table of contents.  An environment is provided for
acknowledgements.

The paradigm of a \LaTeX\ \acronym{UCL} Thesis file is:
\begin{quote}
\begin{verbatim}
\documentclass{ucl_thesis}
\title{<The Thesis Title>}
\author{<The Author>}
\date{<The Year>}
\begin{document}
\maketitle
\begin{abstract}
<Words comprising the abstract.>
\end{abstract}
\begin{acknowledgements}
<Words comprising the acknowledgements.>
\end{acknowledgements}
\tableofcontents
\listoffigures
\listoftables
<... The thesis ...>
\end{document}
\end{verbatim}
\end{quote}

Whilst, the entire thesis could reside in a single file, it is far
more likely that the above will reside in one file with the thesis
content being a set of \verb+\include+ or \verb+input+ calls, perhaps
each chapter being a separate file.

It is strongly advised that Bib\TeX\ be used for handling the
references.  The \verb+\bibliographystyle+ call comes in the document
body, not in the preamble.  The \verb+\bibliography+ call should be
exactly where the bibliography is to appear.  It is strongly
recommended that either \texttt{plain} (essentially Vancouver) or
Harvard citation style are used.  There are a large number of
different implementations of Harvard and derivatives.  I have settled
on one:  \texttt{rnmdt} (mostly because I constructed it!).
\begin{quote}
\begin{verbatim}
\documentclass{ucl_thesis}
\usepackage{rnmdt}
\title{<The Thesis Title>}
\author{<The Author>}
\date{<The Year>}
\begin{document}
\bibliographystyle{rnmdt}
\maketitle
\begin{abstract}
<Words comprising the abstract.>
\end{abstract}
\begin{acknowledgements}
<Words comprising the acknowledgements.>
\end{acknowledgements}
\tableofcontents
\listoffigures
\listoftables
<... The thesis ...>
\bibliography{...}
\end{document}
\end{verbatim}
\end{quote}

This provides a set of \verb+\cite+ macros beyond that specified in
the various books on \LaTeXe.

In the text, one is likely to use the standard \verb+\cite{...}+ quite
a lot.  This produces `(namelist, date)' in the text.  As this is a
name--date style, we often want `namelist (date)'.  For this use
\verb+\citeasnoun{...}+.  \dots\ here represents a label that labels
an item in the Bib\TeX\ database.  Note that whilst
\verb+\cite{label1,label2}+ make sense,
\verb+\citeasnoun{label1,label2}+ is an error.

Possesiveness is sometimes required and this can be a difficulty if
the name is being used as a noun, e.g.\ Winder's (1994) \dots. In this
situation \verb+\citeasposnoun{...}+ can come in handy.  Also
\verb+\citeasplposnoun{...}+ which deals with situation where an s
ends the name and the apostrophe is the last character.

Unfortunately, we sometimes have to do the formatting entirely by
hand.  For example, Winder (1994; 1995; Wernick~\& Winder, 1994)
cannot be handled by any of the above so use:

\begin{quote}
  \verb+\unfcite+ --- produce the `name, date' with no parentheses.\\
  \verb+\unfshcite+ --- produce only `date' with no parentheses.
\end{quote}

These should, however, only be used in this sort of case, they are not
intended for general use.

Finally, there are some circumstances where \verb+\cite+ just cannot
be used without some difficulty.  A citation within a \verb+\caption+
is an example.  In such circumstances, please ensure that you protect
the macro, e.g.\ \verb+\caption{Problem \protect\cite{Winder1994}}+.

\section{On Typefaces}

The department uses Helvetica (or Ariel) and Times (or Times New
Roman) as its standard typefaces and these are the defaults for all
the \acronym{UCL} classes.  People clearly can but
\textbf{\emph{should not}} use different typefaces.

\section{UCL Exam Papers}

This class is used for constructing exam papers according to the rules
laid down by Registrar's Division.  Answers can be included in the
same file and printed or not, to produce the final exam paper.

An exam paper is a 12pt Times-Roman, two-sided, one-column,
space-and-a-half, A4 article comprising questions, subquestions and
subsubquestions each one of which may have an answer and a mark
associated with it.

The cover sheet of an examination paper is provided centrally and
details the usual material: heading, title, course, date, time,
copyright, etc.  It is treated as unnumbered material.  Thus, the exam
paper starts on page~1 with the rubric for the paper.

Style requires that marks should be put for each question or sub-part
thereof with a total at the end.  There is no separate macro for doing
totals, so these must be added manually.

Model answers can (indeed should) be provided within the same file as
the questions.  It is usually best to place the answer directly below
the (sub-)$^2_0$question to which it refers but there is no
requirement that this is so.

All features of an exam paper are provided as environments.  The
source will look something like:
\begin{quote}
\begin{verbatim}
  \begin{rubric}
    ...
  \end{rubric}
  \begin{question}
  \begin{subquestion}
    ...
  \marks{10}
  \begin{answer}
    ...
  \end{answer}
  \end{subquestion}
  \begin{subquestion}
    ...
  \marks{15}
  \begin{answer}
    ...
  \end{answer}
  \marks{Total 25}
  \end{subquestion}
  \end{question}
  \begin{question}
    ...
  \end{question}
\end{verbatim}
\end{quote}
The \verb+\end{...}\begin{...}+ combination happens so often in
questions, subquestions and subsubquestions that a shorthand is
available,
\verb+\further{...}+.  Thus a real paper might look like:
\begin{quote}
\begin{verbatim}
  \begin{rubric}
    ...
  \end{rubric}
  \begin{question}
  \begin{subquestion}
    ...
  \marks{10}
  \begin{answer}
    ...
  \end{answer}
  \further{subquestion}
    ...
  \marks{15}
  \begin{answer}
    ...
  \end{answer}
  \marks{Total 25}
  \end{subquestion}
  \further{question}
    ...
  \end{question}
\end{verbatim}
\end{quote}

The decision of whether to print the answers (which are typeset in the
associated sans serif typeface --- Helvetica when using Times-Roman,
which is what should be used) or not rests with the Boolean
\verb+\ifanswersrequired+, whose default is false --- NB the use of a
primitive \TeX\ Boolean here.  To print answers, the call
\verb+\answersrequiredtrue+ should appear somewhere at the start of
the file, usually in the preamble.

\end{document}
