%!TEX root = ../thesis.tex
% Acknowledgements

\pdfbookmark[1]{Acknowledgements}{Acknowledgements} % Bookmark name visible in a PDF viewer

\begin{flushright}{\slshape    
Es ist nicht genug, zu wissen, man muß auch anwenden;\\
es ist nicht genug, zu wollen, man muß auch tun.
} \\ \medskip
--- Johann Wolfgang von Goethe
\end{flushright}

\bigskip

%----------------------------------------------------------------------------------------

\begingroup

\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax

\chapter*{Acknowledgements} % Acknowledgements section text

\noindent 
Firstly, I would like to express my immense gratitude to my supervisors Claudia Wheeler-Kingshott and Daniel Alexander who gave me the opportunity to study with two of the world-leading research groups. Both have been best supervisors anyone could wish for. I'm very grateful for all the guidance and help I received from them throughout my PhD and I consider myself both very lucky and very honoured, being their student.\\
\noindent
I'm very gratefully towards the International Spinal Research Trust who funded this PhD project.\\ \medskip
\noindent 
I also would like to thank Dr. Patrick Freund, not only for the fruitful collaboration, but also for being a friend and motivator. I also wish to express my sincere thanks to  David Thomas,  Olga Ciccarelli and Carolina Kachramanoglou for providing me with data to work with after our 1.5T scanner had broken down and the new 3T scanner was not installed yet. Many thanks also to Tim Dyrby and Henrik Lundell for inviting me to Denmark and giving me the opportunity to test my protocols on their scanner and samples.\\

\noindent
I would like to thank my friend Laura Panagiotaki for her constant support and valuable help in all the aspects of PhD life. I would also like to thank the both doctors Nedjati-Gilani for all their advice and support.
Many thanks also to Bhavana Solanky and Becky Samson for being great office mates, having been often the only human contacts to the outside world, particularly when another ISMRM deadline was near. \\

\noindent
I also would like to thank all the people in the Microstructure Imaging Group and the Department of Neuroinflammation all the support, but also for being great company in and out of the office. \\

\noindent
A very special thanks goes to my dear friends Franziska Teichert and Frank Penczek for being always there for me for a BBQ or the occasional German fire punch. 
Finally, I would like to express my deepest gratitude to my loving parents Birgit and J\"urgen. I'm infinitely grateful for their encouragement and continuous support for all my life. Last, but not least, I thank Chrystele for joining me on all the way since I arrived in England and for being such a wonderful and inspiring influence on my life ever since.

\endgroup