% Publications - a page listing research articles written using content in the thesis

\pdfbookmark[1]{Publications}{Publications} % Bookmark name visible in a PDF viewer

\chapter*{Publications} % Publications page text

Some ideas and figures have appeared previously in the following publications:

\bigskip
\subsection*{Journal papers}
\begin{itemize}
\item Freund, P., Schneider, T., Nagy, Z., Hutton, C., Weiskopf, N., Friston, K., Wheeler-Kingshott, C. A. M. Thompson, A. (in press). Degeneration of the injured cervical cord is associated with remote changes in corticospinal tract integrity and upper limb impairment. \emph{PLoS One}.
\item Panagiotaki, E., Schneider, T., Siow, B., Hall, M. G., Lythgoe, M. F., \& Alexander, D. C. (2012). Compartment models of the diffusion MR signal in brain white matter: a taxonomy and comparison.. Neuroimage, 59(3), 2241-2254. 
\item Wheeler-Kingshott, C. A. M., Ciccarelli, O., Schneider, T., Alexander, D. C., \& Cercignani, M. (2012). A new approach to structural integrity assessment based on axial and radial diffusivities. \emph{Functional Neurology, 27(2), 85-90}.
\item Zhang, H., Schneider, T., Wheeler-Kingshott, C. A., \& Alexander, D. C. (2012). NODDI: Practical in vivo neurite orientation dispersion and density imaging of the human brain. \emph{NeuroImage, 61(4), 1000-1016}.
\end{itemize}
	
\subsection*{Journal papers in preparation}
\begin{itemize}
 \item Ciccarelli, O., Thomas, D. L., De Vita, E., Wheeler-Kingshott, C. A. M., Schneider, T., Kachramanoglou, C., Toosy, A. T., \& Thompson, A. J. Spinal cord spectroscopy, tractography and q-space MRI in a case of NMO spectrum disorder.
\end{itemize}
\subsection*{Conference papers}
\begin{itemize}
 \item Schneider, T., Wheeler-Kingshott, C. A. M., \& Alexander, D. C. (2010). In-vivo estimates of axonal characteristics using optimized diffusion MRI protocols for single fibre orientation. \emph{13th International Conference on Medical Image Computing and Computer-Assisted Intervention (MICCAI2010)}  
\end{itemize}
\subsection*{Conference abstracts}
\begin{itemize}
\item Schneider, T., Wheeler-Kingshott, C. A. M., \& Alexander, D. C. (2012). Mapping the axon diameter index in the corpus callosum is clinically feasible.\emph{ In Proc. Intl. Soc. Mag. Reson. Med Vol. 20}. Melbourne, Australia.\\ \emph{(selected for oral presentation)}
\item Schneider, T., Ciccarelli, O., Kachramanoglou, C., Thomas, D. L., \& Wheeler-Kingshott, C. A. M. (2011). Reliability of tract-specific q-space imaging metrics in healthy spinal cord. In Proc. \emph{Intl. Soc. Mag. Reson. Med Vol. 19 (pp. 680)}. Montreal, Canada.\\
\emph{(selected for oral presentation)}
\item Schneider, T., Thomas, D. L., Kachramanoglou, C., Ciccarelli, O., Alexander, D. C., \& Wheeler-Kingshott, C. (2011). Fuzzy partial volume correction of spinal cord DTI parameters. \emph{In Proc. Intl. Soc. Mag. Reson. Med Vol. 19 (pp. 4556).} Montreal, Canada.
\item	Schneider, T., Nagy, Z., Wheeler-Kingshott, C. A. M., Thomson, A. J., \& Freund, P. (2011). Diffusion tensor imaging detects axonal degeneration and its extent is associated with disability in chronic spinal cord injury. \emph{In Proc. Intl. Soc. Mag. Reson. Med Vol. 19 (pp. 2353)}. Montreal, Canada.
\item	Schneider, T., Alexander, D. C., \& Wheeler-Kingshott, C. A. M. (2010). Optimized diffusion MRI protocols for estimating axon diameter with known fibre orientation. \emph{In Proc. Intl. Soc. Mag. Reson. Med Vol. 18 (pp. 1561)}. Stockholm, Sweden.
\item	Schneider, T., Alexander, D. C., \& Wheeler-Kingshott, C. A. M. (2009). Preliminary investigation of position dependency of radial diffusivity in the cervical spinal cord. \emph{In Proc. Intl. Soc. Mag. Reson. Med Vol. 17 (pp. 3191)}. Honolulu, Hawaii.
\end{itemize}

% \subsection*{Invited talks}
% \begin{itemize}
%  \item ``Optimising Diffusion Pulse Sequences for Investigation of Spinal Cord Microstructure'', Departmental Seminar, Danish Research Centre for Magnetic Resonance, Copenhagen, August 2009
%  \item ``Spinal Cord Diffusion MRI'', CMIC Seminar, Centre for Medical Image Computing, UCL, London, January 2010
%  \item ``Imaging microstructure in the spinal cord with diffusion MRI'', Imaging \& Biophysics Unit Seminar Series, Institute of Child Health, UCL, London, November 2010   
% \end{itemize}
